---
title: "Starting with data"
order: 14
fig-align: center
fig-cap-location: bottom
engine: knitr
execute:
  echo: true
  code-tools: true
description: "In this lesson we begin to work with data. We start by loading data from a file into a dataframe, then begin to work with the dataframe."
---

### Loading survey data

Today we'll work with animal species data, stored as comma-separated values (CSV) in a `.csv` file in the following format:

| Column | Description | 
|---------------|--------------------------------------------|
| record\_id       | Unique id for the observation                | 
| month           | month of observation                         | 
| day             | day of observation                           | 
| year            | year of observation                          | 
| plot\_id         | ID of a particular experimental plot of land | 
| species\_id      | 2-letter code                                | 
| sex             | sex of animal ("M", "F")                     | 
| hindfoot\_length | length of the hindfoot in mm                 | 
| weight          | weight of the animal in grams                | 
| genus           | genus of animal                              | 
| species         | species of animal                            | 
| taxon           | e.g. Rodent, Reptile, Bird, Rabbit           | 
| plot\_type       | type of plot                                 | 

---

**Downloading the data**

We first need to download the data using the `download.file()` function.

Checking the documentation with `?download.file()`, we see that the function takes two important options

- `url` -- a character string naming the URL of a resource to be downloaded
- `destfile` -- a character string with the file path where the file will be saved

We're going to save this to our `data_raw` directory that we created last session.

```{r}
#| eval: true
#| echo: true

download.file(url = "https://ndownloader.figshare.com/files/2292169",
              destfile = "~/data-carpentry/data_raw/portal_data_joined.csv")
```

---

**Reading the data into R**

Now we can load the data from the computer into an object in R. We'll do this with the `read_csv()` function from the `tidyverse` package.

::: {.callout-note}

# Base R function and packages

Until now we've used base `R` functions that are built into `R`. Packages provide acccess to additional functions beyond those that come with `R`. 

Here we use functions from the package `tidyverse`.

First, let's install `tidyverse` (if you haven't already).

```{r}
#| eval: false
#| echo: true
install.packages("tidyverse")
```

:::

---

Let's load the `tidyverse` package.

```{r}
#| eval: true
#| echo: true
library(tidyverse)
```

---

Now we can read the data from the file into a dataframe object, `surveys`.

```{r}
#| eval: true
#| echo: true
surveys <- read_csv("~/data-carpentry/data_raw/portal_data_joined.csv")
```

Notice the data types are set for each column of `surveys`, even though we didn't set them. This is because `read_csv` will look at the first 1000 rows of each column and guess the data type.

:::{.callout-note}

`read_csv()` assumes that fields are delineated by commas. However, this may not always be the case. There are several other functions to read in data in a tabular format. The most generalizable function would be `read_delim()` which allows for the specification of column / field delimiters.

:::

---

Let's view the contents of the first 6 rows of survey by using the `head()` function.

```{r}
#| eval: true
#| echo: true
head(surveys)
```

We can change the number of rows we view by passing the option `n` to the function.

```{r}
#| eval: true
#| echo: true
head(surveys, n=7)
```

---

Now let's look at the last 7 lines using `tail()`, which works in a similar fashion.
```{r}
tail(surveys, n=7)
```

---

For a more complete view of `surveys` we can use the `view()` command.

```{r}
#| eval: false
#| echo: true
view(surveys)
```

:::{.callout-note}

There are two functions for viewing which are case-sensitive. Using `view()` with a
lowercase 'v' is part of `tidyverse`, whereas using `View()` with an uppercase 'V' is
loaded through base `R` in the `utils` package.

:::

---

### What are dataframes?

Dataframes are a VERY common data structure for tabular data.

- columns are vectors of the same length
- each column must contain a single data type (characters, integers, etc.)

An example of this is shown in the figure below.

::: {#fig-dataframe}

![](/images/data-frame.svg)

A dataframe comprising a numeric, a character, and a logical vector. [1]
:::

---

We can also see this when inspecting the **str**ucture of the dataframe using the function `str()`.

```{r}
#| eval: true
#| echo: true
str(surveys)
```

---

:::{.callout-note}
When we loaded the data into `R`, it got stored as an object of class `tibble`, which is a special kind of data frame (the difference is not important for our purposes, but you can learn more about tibbles [here](https://tibble.tidyverse.org/)).
:::

---

### Inspecting dataframes

There are several functions that can be useful in checking content, structure, and other characteristics fo the dataframe. Here is a incomplete list.

| Characteristic | Function | Descritption |
|----------------|----------|--------------|
| Size | `dim(suverys)` | returns a vector with the number of rows in the first element, and the number of columns as the second element (the **dim**ensions of the object) |
|      | `nrow(surveys)` | returns the number of rows |
|      | `ncol(surveys)` | returns the number of columns |
| Content | `head(surveys)` | shows the first n rows, 6 by default |
|         | `tail(surveys)` | shows the last n rows, 6 by default |
| Names   | `names(surveys)` | returns the column names (synonym of `colnames()` for `data.frame` objects) |
|         | `rownames(surveys)` | returns the row names |
| Summary | `str(surveys)` | structure of the object and information about the class, length and content of  each column |
|         | `summary(surveys)` | summary statistics for each column |

: {tbl-colwidths="[20,20,60]"}

Take some time to try these out by copy/pasting these into your console.

---

#### Challenge

:::{.panel-input}

_**Q&A**_: Based on the output of `str(surveys)`, can you answer the following questions?

- What is the class of the object `surveys`?
- How many rows and how many columns are in this object?

::: {.callout-note collapse="true" icon="false"}
### _Click here for the answer_  

```{r}
#| eval: true
#| echo: true
str(surveys)
```

Class: dataframe
Rows: 34786
Columns: 13

:::

:::

---

### Indexing and subsetting dataframes

Dataframes are 2 dimensional data types. The location of each item in a dataframe can be described by the indices of these two dimensions (row and column indices) -- where row number comes first, followed by column number.

::: {.callout-important}
Different ways of specifying these coordinates lead to results with different classes.
:::

---

#### Extracting specific values

`dataframe[row_index, column_index]`

First row, first column
```{r}
#| eval: true
#| echo: true
surveys[1,1]
```

---

First row, sixth column
```{r}
#| eval: true
#| echo: true
surveys[1,6]
```

---

#### Extracting entire rows or columns  

`datarame[row_index, ]` -- specific row, all columns  
`datarame[, col_index]` -- all rows, specific column

First row, all columns
```{r}
#| eval: true
#| echo: true
surveys[1, ]
```

---

All rows, first column
```{r}
#| eval: true
#| echo: true
surveys[, 1]
```

---

Another way to get a specific column, all rows. _Note, there is no comma used._
```{r}
#| eval: true
#| echo: true
surveys[1]
```

---

#### Multiple rows or columns -- use vectors to specify indices

`data_frame[c(row_index, row_index, rowindex), c(column_index, column_index)]`

The first three rows of the fifth and sixth columns
```{r}
#| eval: true
#| echo: true
surveys[c(1, 2, 3), c(5, 6)]
```

---

Another way, use the `:` to select a range of indces
```{r}
#| eval: true
#| echo: true
surveys[1:3, 5:6]
```

---

#### Returning vectors -- use `[[]]`

As we've seen thus far when working with tibbles, subsetting with `[]` returns a dataframe.

If we want vectors returned we can use `[[]]`

The first column as a vector
```{r}
#| eval: false
#| echo: true
surveys[[1]]
```

---

The first value in our dataframe.
```{r}
surveys[[1,1]]
```

---

#### Excluding indices -- use `-`

All rows, all but the first column
```{r}
#| eval: true
#| echo: true
surveys[, -1]
```

---

The equivalent of head (first 6 rows, all columns -- or all but the 7th through max row number).
```{r}
#| eval: true
#| echo: true
surveys[-(7:nrow(surveys)), ]
```

---

#### Calling column names directly

Dataframes can also be subet by calling the names of the columns directly.

Using column names to return dataframes.
```{r}
#| eval: false
#| echo: true
surveys["species_id"]
surveys[, "species_id"]
```

---

Using column names to return vectors.
```{r}
#| eval: false
#| echo: true
surveys[["species_id"]]
```

---

We can also use the `$` operator to call column names instead of `[[]]`.
```{r}
#| eval: false
#| echo: true
surveys$species_id
```

---

#### Challenge

:::{.panel-input}

_**Q&A**_: Write code that does the following:

1. Create a dataframe (`surveys_200`) containing only the data in row 200 of the `surveys` dataset.
2. Notice how `nrow()` gave you the number of rows in a dataframe?

  - Use that number to pull out just the last row of the `surveys` dataset.
  - Compare that with the output of using `tail()` to make sure it meets expectations.
  - Pull out the last row using `nrow()` instead of the row number.
  - Create a new dataframe (`surveys_last`) from that last row.

3. Use `nrow()` to extract the row that is in the middle of the dataframe. Store the content of this row in an object named `surveys_middle`.

4. Combine `nrow()` with the `-` notation above to reproduce the behavior of `head(surveys)`, keeping just the first through sixth of the `surveys` dataset.

::: {.callout-note collapse="true" icon="false"}
### _Click here for the answer_  

```{r}
#| echo: true
#| eval: true

## 1.
surveys_200 <- surveys[200, ]
## 2.
# Saving `n_rows` to improve readability and reduce duplication
n_rows <- nrow(surveys)
surveys_last <- surveys[n_rows, ]
## 3.
surveys_middle <- surveys[n_rows / 2, ]
## 4.
surveys_head <- surveys[-(7:n_rows), ]

```

:::

:::

---

### Factors

Using `str(surveys)` showed us that several of the columns consist of categorical data, meaning there are a limited number of values (categories). This is contrast to some of the other columns that have a large range of possible number values.

`Factor` -- a special class for working with categorical data  

- Once created, factors can only contain a _pre-defined_ set of values, know as `levels`
- Stored as integers associated with labels
- Can be ordered or unordered
- Are treated as integer vectors

We can convert columns to the `factor` data type using the `factor()` function.
```{r}
#| eval: true
#| echo: true
surveys$sex <- factor(surveys$sex)
```

Let's check the conversion with `summary()`.
```{r}
#| eval: true
#| echo: true
summary(surveys$sex)
```

---

By default, `R` always sorts levels in alphabetical order. For instance, if you have a factor with 2 levels, like below.
```{r}
#| eval: true
#| echo: true
sex <- factor(c("male", "female", "female", "male"))
```

`R` will assign 

- `1` to the level `"female"`
- `2` to the level `"male"` 

You can see this by using the function `levels()`.
```{r}
#| eval: true
#| echo: true
levels(sex)
```

You can find the number of level using `nlevels()`

```{r}
#| eval: true
#| echo: true
nlevels(sex)
```

---

Sometimes, the order of the factors does not matter, other times you might want to specify the order because it is meaningful (e.g., "low", "medium", "high"), it improves your visualization, or it is required by a particular type of analysis. 


Here, one way to reorder our levels in the `sex` vector would be:

```{r}
#| eval: true
#| echo: true
sex # current order
sex <- factor(sex, levels = c("male", "female"))
sex # after re-ordering
```

---

While factors are represented by integers (1, 2, 3), they are more informative because they are self-describing: `"female"`, `"male"` is more descriptive than `1`, `2`. 

Factors have this information built in, which is helpful when there are many levels (like the species names in our example dataset).

---

#### Challenge

:::{.panel-input}

_**Q&A**_: What code would you write to do the following?

1. Change the columns `taxa` and `genus` in the `surveys` dataframe into a factor.
2. Using the functions you learned before, can you find out...

  - How many rabbits were observed?
  - How many different genera are in the `genus` column?

::: {.callout-note collapse="true" icon="false"}
### _Click here for the answer_  
  
```{r}
#| eval: true
#| echo: true
num_char <- c(1, 2, 3, "a")
surveys$taxa <- factor(surveys$taxa)
surveys$genus <- factor(surveys$genus)
summary(surveys)
nlevels(surveys$genus)

## * how many genera: There are 26 unique genera in the `genus` column.
## * how many rabbts: There are 75 rabbits in the `taxa` column.
```
:::

:::

---

### Converting Factors

To convert a factor to a character vector, you can use the `as.character()` function.

```{r}
#| eval: true
#| echo: true
as.character(sex)
```

---

While factors are an efficient way to represent data (because each unique character value is represented once), there may be times when you want to convert factors to numeric vectors, for example, performing simple mathematical operations. 

For this, you cannot use `as.numeric()`, since this will just return the index values of the factor, not the levels. To avoid this, we conver factors --> characters --> numbers.

Another method is to use the `levels()` function.

Here's an example.
```{r}
#| echo: true
#| eval: true
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)               # Wrong! And there is no warning...
as.numeric(as.character(year_fct)) # Works...
as.numeric(levels(year_fct))[year_fct]    # The recommended way.
```

---

Let's break down the `levels()` approach.

There are 3 steps that occur:

1. We obtain all the factor levels using `levels(year_fct)`
2. We convert these levels to numeric values using `as.numeric(levels(year_fct))`
3. We access these numeric values using the underlying integers of the vector `year_fct` inside the square brackets.

---

### Renaming factors

When your data are stored as factors, you can use the `plot()` function to get a quick glance at the number of observations represented by each factor level. 

Let's look at the number of males and females captured over the course of the experiment.
```{r}
#| eval: true
#| echo: true
## bar plot of the number of females and males captured during the experiment
plot(surveys$sex)

```

---

In the plot above, we're missing the individuals for which sex information was not recorded. To show them in the plot we can convert the missing values into a factor level with `addNA()`. We'll need to also give the factor a label.

Note: We are going to make a copy of the `sex` column so that we are not modigying the working copy of the dataframe.

```{r}
#| eval: true
#| echo: true
sex <- surveys$sex #create a copy
levels(sex) #look at the levels
sex <- addNA(sex) #addNA as a level
levels(sex) #check the levels again
head(sex) #look at the data
levels(sex)[3] <- "undetermined" #assign the value "undetermined" to the third, by index, level
levels(sex) #check the levels again
head(sex) #look at the dataset again
```

Now we can plot the data again, using `plot(sex)`.
```{r}
#| eval: true
#| echo: true
plot(sex)
```

#### Challenge
:::{.panel-input}

_**Q&A**_: Write the code to do the following:

- Rename "F" and "M" to "female" and "male" respectively.
- Now that we have renamed the factor level to "undetermined", can you create the bar plot such that "undetermined" is first (before "female")?

::: {.callout-note collapse="true" icon="false"}
### _Click here for the answer_  
  
```{r}
#| eval: true
#| echo: true
levels(sex)[1:2] <- c("female", "male")
sex <- factor(sex, levels = c("undetermined", "female", "male"))
plot(sex)
```

:::

:::

:::{.panel-input}

_**Q&A**_: We have seen how dataframes are created when using `read_csv()`, but they can also be created by hand with the `data.frame()` function.  There are a few mistakes in this hand-crafted `data.frame`. Can you spot and fix them? Don't hesitate to experiment!

```{r}
#| eval: false
#| echo: true

animal_data <- data.frame(
          animal = c(dog, cat, sea cucumber, sea urchin),
          feel = c("furry", "squishy", "spiny"),
          weight = c(45, 8 1.1, 0.8)
          )
  
```

::: {.callout-note collapse="true" icon="false"}
### _Click here for the answer_  
  
1. The animals listed have no quotes around them.
2. There are not the same number of items in each vector.
3. There is a missing comma in the weight vector.

:::

:::

:::{.panel-input}

_**Q&A**_: Can you predict the class for each of the columns in the following example?

- Are they what you expected?  Why? Why not?
- What would you need to change to ensure that each column had the accurate data type?

```{r}
#| eval: false
#| echo: true
country_climate <- data.frame(
       country = c("Canada", "Panama", "South Africa", "Australia"),
       climate = c("cold", "hot", "temperate", "hot/temperate"),
       temperature = c(10, 30, 18, "15"),
       northern_hemisphere = c(TRUE, TRUE, FALSE, "FALSE"),
       has_kangaroo = c(FALSE, FALSE, FALSE, 1)
       )
```

::: {.callout-note collapse="true" icon="false"}
### _Click here for the answer_   

- `country` --> character  
- `climate` --> character  
- `temperature` --> character  
- `northern_hemisphere` --> character  
- `has_kangaroo` --> numeric  

:::

:::

---

The automatic conversion of data type is sometimes a blessing, sometimes an
annoyance. 

Be aware that it exists, learn the rules, and double check that data
you import in `R` are of the correct type within your data frame. If not, use it
to your advantage to detect mistakes that might have been introduced during data
entry (for instance, a letter in a column that should only contain numbers).

### Formatting dates

A common challenge for `R` users is converting date and time information into an analysis-suitable format.

One way to start date information is to store each component (day, month, year) in a separate column. We can see that`surveys` takes this approach.
```{r}
#| eval: true
#| echo: true
str(surveys)
```

We'll use the `ymd()` function from the `lubridate` packages (installed with `tidyverse`).

Load the package.
```{r}
#| eval: true
#| echo: true
library(lubridate)
```

---

Let's create a data object and inspect the structure.
```{r}
#| eval: true
#| echo: true
my_date <- ymd("2015-01-01")
str(my_date)
```

---

Now let's paste the year, month, and day separately - we get the same result.
```{r}
#| eval: true
#| echo: true
# sep indicates the character to use to separate each component
my_date <- ymd(paste("2015", "1", "1", sep = "-")) 
str(my_date)
```

---

Now we apply this function to the surveys dataset. Create a character vector from the `year`, `month`, and `day` columns of `surveys` using `paste()`

```{r}
#| eval: true
#| echo: true
head(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
```

This character vector can be used as the argument for `ymd()`:

```{r}
#| eval: true
#| echo: true
head(ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-")))
```

There is a warning telling us that some dates could not be parsed (understood)
by the `ymd()` function. For these dates, the function has returned `NA`, which
means they are treated as missing values -- we will deal with this problem later. 

---

First we add the resulting `Date` vector to the `surveys` dataframe as a new column called `date`.

```{r, purl=FALSE}
surveys$date <- ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
str(surveys) # notice the new column, with 'date' as the class
```

---

Let's make sure everything worked correctly. One way to inspect the new column is to use `summary()`.

```{r, results=TRUE, purl=FALSE}
summary(surveys$date)
```

---

So why were some dates not parsed?

We can use the functions we saw previously to deal with missing data to identify
the rows in our data frame that are failing. 

If we combine them with what we learned about subsetting data frames earlier, we can extract the columns "year, "month", "day" from the records that have `NA` in our new column `date`. We will also use `head()` so we don't clutter the output.

```{r}
#| eval: true
#| echo: true
missing_dates <- surveys[is.na(surveys$date), c("year", "month", "day")]

head(missing_dates)
```

Looking at the dates that were not converted, it seems that they don't actualy exist -- September and April only have 30 days, not 31 days as it is specified in our dataset.

There are several ways you could deal with situation:

- If you have access to the raw data (e.g., field sheets) or supporting information (e.g., field trip reports/logs), check them and ensure the electronic database matches the information in the original data source.
- If you are able to contact the person responsible for collecting the data, you could refer to them and ask for clarification.
- You could also check the rest of the dataset for clues about the correct value for the erroneous dates.
- If your project has guidelines on how to correct this sort of errors, refer to them and apply any recommendations.
- If it is not possible to ascertain the correct value for these observations, you may want to leave them as missing data.

Regardless of the option you choose, it is important that you document the error and the corrections (if any) that you apply to your data.

---

### Citations

1. Data Analysis and visualization in R for ecologists. Data Analysis and Visualisation in R for Ecologists: Data Analysis and Visualization in R for Ecologists. <https://datacarpentry.org/R-ecology-lesson/index.html>